---
title: "Accuracy of discrete formulas for life-table shape measure"
author: "Jonas Schöley"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{accuracy}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r set-options, echo=FALSE, cache=FALSE}
options(width = 100)
knitr::opts_chunk$set(tidy = FALSE, size = "small")
```

In order to assess the accuracy of life-table summary measures over varying levels of data aggregation I distinguish between 3 levels of age discreteness:

- Level 0: Calculation of summary measure directly from individual level survival times
- Level 1: Calculation of summary measure from complete (single year) life-table
- Level 2: Calculation of summary measure from abridged (10 year) life-table

The better the discretization of the summary measure the closer its level 1 and 2 values are to the value calculated directly from level 0 data.

```{r init}
library(pash)
library(tidyverse)

radix = 10000

# create pash objects
lt_lvl1 <- Inputlx(x = australia_1y$x, lx = australia_1y$lx,
                   nax = australia_1y$nax, nx = australia_1y$nx, last_open = TRUE)
lt_lvl2 <- Inputlx(x = australia_10y$x, lx = australia_10y$lx,
                   nax = australia_10y$nax, nx = australia_10y$nx, last_open = TRUE)
```

```{r e0}
# not a shape measure, just as a test, should be exactly equal
e0 <- c(
  lvl0 = mean(australia_surv_times),
  lvl1 = GetPace(lt_lvl1, type = "e0"),
  lvl2 = GetPace(lt_lvl1, type = "e0")
)
e0
```

Gini coefficient
----------------

Formulation of individual level Gini coefficient: Damgaard, C., & Weiner, J. (2000). Describing inequality in plant size or fecundity. Ecology, 81(4), 1139-1142.

```{r gini}
X = data.frame(i = 1:length(australia_surv_times), x = sort(australia_surv_times))
G_lvl0 = sum((2*X$i-nrow(X)-1)*X$x)/(nrow(X)^2*mean(australia_surv_times))

gini <- c(
  lvl0 = G_lvl0,
  lvl1 = GetShape(lt_lvl1, type = "gini", harmonized = FALSE),
  lvl2 = GetShape(lt_lvl2, type = "gini", harmonized = FALSE)
)

gini
```

Coefficient of variation
------------------------

```{r cv}
cv <- c(
  lvl0 = sd(australia_surv_times) / mean(australia_surv_times),
  lvl1 = GetShape(lt_lvl1, type = "cv", harmonized = FALSE),
  lvl2 = GetShape(lt_lvl2, type = "cv", harmonized = FALSE)
)

cv
```

Life-table entropy
------------------

```{r entropy}
eDaggerx = vector(mode = "double", length = length(australia_surv_times))
for (i in seq_along(australia_surv_times)) {
  if (i < length(australia_surv_times)) {
    eDaggerx[i] =
      mean(australia_surv_times[seq(i+1, length(australia_surv_times))]) -
      australia_surv_times[i]
  }
}

entropy <- c(
  lvl0 = sum(eDaggerx)/radix/mean(australia_surv_times),
  lvl1 = GetShape(lt_lvl1, type = "entropy", harmonized = FALSE),
  lvl2 = GetShape(lt_lvl2, type = "entropy", harmonized = FALSE)
)

entropy
```

Probability to survive to the mean age at death
-----------------------------------------------

```{r psmad}
psmad <- c(
  lvl0 = sum(australia_surv_times > mean(australia_surv_times)) / radix,
  lvl1 = GetShape(lt_lvl1, type = "psmad", harmonized = FALSE),
  lvl2 = GetShape(lt_lvl2, type = "psmad", harmonized = FALSE)
)

psmad
```

Life-expectancy ratio
---------------------

```{r ler}
# remaining life-expectancy at mean age of death (expressed as share of total life-expectancy)
ler <- c(
  lvl0 = (mean(australia_surv_times[australia_surv_times > mean(australia_surv_times)])-mean(australia_surv_times))/mean(australia_surv_times),
  lvl1 = GetShape(lt_lvl1, type = "ler", harmonized = FALSE),
  lvl2 = GetShape(lt_lvl2, type = "ler", harmonized = FALSE)
)

ler
```

Results
-------

```{r results}
data.frame(cv, gini, entropy, ler, psmad) %>%
  mutate(lvl = rownames(.)) %>%
  gather(key = measure, value = value, -lvl) %>%
  group_by(measure) %>%
  do({
    rbind(
      data.frame(lvl = 1, error = (.$value[2]-.$value[1])/.$value[1]),
      data.frame(lvl = 2, error = (.$value[3]-.$value[1])/.$value[1])
    )
  }) %>%
  ggplot() +
  geom_col(aes(x = interaction(lvl, measure), y = error, fill = measure)) +
  scale_x_discrete("", labels = c("CV-1Y.", "CV-10Y", "Entropy-1Y", "Entropy-10Y",
                                  "Gini-1Y", "Gini-10Y", "LER-1Y", "LER-10Y",
                                  "PSMAD-1Y", "PSMAD-10Y")) +
  scale_y_continuous(labels = scales::percent_format(), limits = c(-0.05, 0.05),
                     breaks = seq(-0.05, 0.05, 0.01)) +
  theme_minimal() +
  theme(panel.grid.minor = element_blank(), panel.grid.major.y = element_blank()) +
  guides(fill = FALSE) +
  coord_flip()
```
