---
title: "pash: Pace-Shape Analysis for Life-tables"
author: "Jonas Schoeley, Marius Pascariu, Francisco Villavicencio, Maciej Danko"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{pash}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

What's it for?
--------------

*pash* lets you perform pace-shape analysis[^1][^2][^3] on life-tables. Calculate a wide array of pace and shape metrics and use them to standardize and compare your data.

*pash* is based on a powerful implementation of single-decrement life-tables. Allowing for various input formats, a choice of sensible defaults and extensive error-checking routines make the construction of life-tables extremely easy with *pash* while experts can choose among different life-table estimation methods.

- Step 1: Create a pace-shape object from life table columns using `Inputlx()` or `Inputmx()`.
- Step 2: Using the pace-shape object as argument, calculate pace measures with `GetPace()`, calculate shape measures with `GetShape()`, standardize a life table by pace and shape with `StandardizeLT()`, re-base your life table to a different age with `RebaseLT()`.

[^1]: Baudisch, Annette. 2011. "The Pace and Shape of Ageing." Methods in Ecology and Evolution 2 (4): 375–82. doi:10.1111/j.2041-210X.2010.00087.x.
[^2]: Wrycza, Tomasz, and Annette Baudisch. 2014. "The Pace of Aging: Intrinsic Time Scales in Demography." Demographic Research 30 (1): 1571–90. doi:10.4054/DemRes.2014.30.57.
[^3]: Wrycza, Tomasz F., Trifon I. Missov, and Annette Baudisch. 2015. "Quantifying the Shape of Aging."" PLOS ONE 10 (3): 1–18. doi:10.1371/journal.pone.0119163.

How to install?
---------------

You can install `pash` by running:

```{r eval=FALSE}
install.packages("devtools")
devtools::install_github("jschoeley/pash")
```

Examples
--------

```{r}
library(pash)
```

### Create a life-table from mortality rates

Let's say we have a vector of mortality rates by age (Sweden, females, 1940--1944) and would like to calculate a life-table from it...

```{r}
swe <- subset(sweden5x5, sex == "female" & period == "1940-1944")[c("x", "nmx")]
swe
```

...we use the `Inputnmx()` function and specify the last age group as open, done.

```{r}
Inputnmx(x = swe$x, nmx = swe$nmx, last_open = TRUE)
```

We can change the details of the life-table calculation.

```{r}
Inputnmx(x = swe$x, nmx = swe$nmx, last_open = TRUE, nax = "cfm")
```

### Create a life-table from a survival curve

Given a vector of age-specific survival probabilities we can easily derive a whole life-table. By default the last age group is assumed to be closed.

```{r}
prestons_lx
Inputlx(x = prestons_lx$x, lx = prestons_lx$lx)
```

### Calculate measures of pace and shape

```{r}
pash = Inputlx(x = prestons_lx$x, lx = prestons_lx$lx)
```

Once a `pash` object has been created using one of the `Input*()` functions pace and shape measures are readily available. By default all available pace measures are returned.

```{r}
GetPace(pash)
```

The age where only 20% of the population is alive:

```{r}
GetPace(pash, type = "qlx", q = 0.2)
```

By default all available shape measures are returned.

```{r}
GetShape(pash)
```

The life-table entropy:

```{r}
GetShape(pash, type = "entropy")
```

### Rebase a life-table

Sometimes life-table shape and pace measures are only calculated for the adult ages. *pash* lets you "re-base" a life-table to a new age.

Create a pash object from a survival curve.

```{r}
pash = Inputlx(x = prestons_lx$x, lx = prestons_lx$lx)
pash
```

Re-base to age 15.

```{r}
pash = RebaseLT(pash, origin = 15)
pash
```

Calculate shape measures on the re-based life-table.

```{r}
GetPace(pash)
GetShape(pash)
```

Re-basing is a non-destructive operation and thus can be reversed:

```{r}
RebaseLT(pash, origin = 0)
```

### Standardize a life-table by pace and shape

Different life-tables can be made comparable in shape by standardizing them by pace.[^3] *pash* allows you to standardize your life-table by different shape measures, total life expectancy being the default:

```{r}
pash = Inputlx(x = prestons_lx$x, lx = prestons_lx$lx)
StandardizeLT(pash)
```

Standardize by median life-span:

```{r}
StandardizeLT(pash, pace = "qlx")
```

Standardize by 0.7 quantile life-span:

```{r}
StandardizeLT(pash, pace = "qlx", q = 0.7)
```

### Using pash within dplyr pipe

Here we use *pash* in conjunction with `dplyr` and `ggplot2` to produce a plot of life expectancy at birth versus life-table entropy for hundreds of Swedish life-tables.

```{r message=FALSE}
library(dplyr)
library(ggplot2)

sweden5x5 %>%
  group_by(sex, period) %>%
    do({
     pash = Inputnmx(x = .$x, nmx = .$nmx)
     data.frame(e0 = GetPace(pash, type = "e0"),
                H  = GetShape(pash, type = "entropy"))
    }) %>%
  ggplot(aes(x = e0, y = H, color = sex)) + geom_point()
```

